# Reproducible Research: Peer Assessment 1
author: "David Parker"  
date: "Friday, August 15, 2014"  
output: html_document  

## Loading and preprocessing the data
```{r}
library(plyr)
library(ggplot2)
library(timeDate)
library(knitr)
opts_knit$set(progress=FALSE)
opts_chunk$set(echo=TRUE, message=FALSE, tidy=TRUE, comment=NA,
               fig.path="figure/", fig.keep="high", fig.width=10, fig.height=6,
               fig.align="center")
activity <- read.csv('activity.csv', colClasses = c('integer', 'Date', 'integer'), 
                     stringsAsFactors = FALSE)
activity$date <- as.POSIXlt(activity$date, tz = '', '%Y%m%d')
```

Display head, tail, and summary stats:  
```{r}
rbind(head(activity,3), tail(activity,3))
summary(activity)
```


## What is mean total number of steps taken per day?
```{r}
actDaily <- ddply(activity, .(date), summarize, 
                  dailySteps = sum(steps, na.rm = TRUE))
```

**Histogram** of *total* daily *steps*:  
```{r}
ggplot(actDaily, aes(x = dailySteps)) + geom_histogram(binwidth = 710)
```

**Mean** daily *steps*:  
```{r}
round(mean(actDaily$dailySteps, na.rm = TRUE), 0)
```

**Median** daily *steps*:  
```{r}
round(median(actDaily$dailySteps, na.rm = TRUE), 0)
```


## What is the average daily activity pattern?  

**Calculate** the *mean* *steps* for each *interval*:  
```{r}
actInterval <- ddply(activity, .(interval), summarize, 
                     meanSteps = round(mean(steps, na.rm = TRUE), 0))
```

**Merge** the *mean* *steps* for each *interval* with the *activity* data frame:  
```{r}
actMerge <- join(x = activity, y = actInterval, by = 'interval')
```

**Plot** of the *5-minute* interval and the *average* number of *steps* taken, *averaged* across all  days:  
```{r}
ggplot(actMerge, aes(x = interval, y = meanSteps)) + geom_line()
```

Five minute *interval* across all days with the **most** *steps*:  
```{r}
actMerge[which.max(actMerge$steps), c('steps','date', 'interval')]
```


## Imputing missing values  

**Total** number of rows with **missing** values is:  
```{r}
nrow(activity) - nrow(na.omit(activity))
```  
This number agrees with the NA's shown in summary from step 1.  

Now to fill in the missing values using the mean of steps for said interval.  
The mean steps is caculated in *actMerge* data frame replacing *NA*'s with average *steps* per *interval*. Simply need to write this data back to the *activity* data frame.  

````{r}
actMerge[which(is.na(actMerge$steps)),]$steps <- 
 actMerge[which(is.na(actMerge$steps)),]$meanSteps
activity <- actMerge[, c('steps', 'date', 'interval')]
```

Recompute the *mean* and *median* daily *steps* with **imputed** values.  
```{r}
actDaily <- ddply(activity, .(date), summarize, 
                  dailySteps = sum(steps, na.rm = TRUE))
```

**Histogram** of *total* daily *steps* using *imputed* values:  
```{r}
ggplot(actDaily, aes(x = dailySteps)) + geom_histogram(binwidth = 710)
```


**Mean** daily *steps*, *imputed* values:  
```{r}
round(mean(actDaily$dailySteps, na.rm = TRUE), 0)
```

**Median** daily *steps*, *imputed* values:  
```{r}
round(median(actDaily$dailySteps, na.rm = TRUE), 0)
```

The impact of imputing missing data was to produce values and plots that were more inline with average steps values. This makes sense due to the fact that I computed mean steps for corresponding intervals for the missing values. The estimates for mean and median and resulting histogram were inline with the averages.  

## Are there differences in activity patterns between weekdays and weekends?  

Create a **wkday** factor variable consisting of either *weekday* or *weekend* values.  
```{r}
activity$wkday <- factor(
  ifelse(isWeekday(activity$date), paste('weekday'), paste('weekend') ))
```

Using dataset with **imputed** values to recalculate the *average steps* taken each day:  
```{r}
actIntDay <- ddply(activity[activity$wkday == 'weekday',], .(interval), summarize, 
                     meanSteps = round(mean(steps, na.rm = TRUE), 0))
actIntEnd <- ddply(activity[activity$wkday == 'weekend',], .(interval), summarize, 
                   meanSteps = round(mean(steps, na.rm = TRUE), 0))
actMergDay <- join(x = activity, y = actIntDay, by = 'interval')
actMergEnd <- join(x = activity, y = actIntEnd, by = 'interval')
actMerge <- 
  rbind(actMergDay[actMergDay$wkday == 'weekday',], actMergEnd[actMergEnd$wkday == 'weekend',])
```  

**Plot** of the 5-minute interval and the average number of steps taken, averaged across all days  using *wkday* factor as the facet, wrapping on rows:  
```{r}
ggplot(actMerge, aes(x = interval, y = meanSteps)) + 
       geom_line() + facet_wrap(~wkday, nrow = 2)
```  
  
The patterns were very similar with the weekend showing a more "relaxed" overall appearance which would seem to be expected.  